---
title: "Quantified self movement analysis"
author: "MLMV"
date: "29 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(ggplot2)
library(rpart)
library(randomForest)
```


***
### Executive summary

Using Fitbit or similar devices, many people like to quantify how much of a particular activity they do. However they rarely quantify *how well* they do it. In this project, our goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants who were asked to perform barbell lifts in 5 different ways correctly and incorrectly, and to build a model that can predict the manner in which they did the exercise.



***
### Data & exploratory analysis
The data for this project comes from http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises.

```{r load_data}
training = read.csv("pml-training.csv", na = c("", "NA"))
testing = read.csv("pml-testing.csv", na = c("", "NA"))
```

Before starting the modeling, we do some data profiling. The data consists of `r dim(training)[2]` variables, with `r dim(training)[1]` observations in a training set and `r dim(testing)[1]` additional observations in a testing set. The outcome variable "classe" is roughly evenly distributed, overall and across users, with some prevalence for classe A.

```{r fig.height = 2.5}
par(mfrow = c(1,2), cex = 0.6)
barplot(table(training$classe), main = "Activity by classe", xlab = "Classe", ylab = "Count")
plot(classe ~ user_name, data = training, main = "Activity by user", xlab = "User", ylab = "Classe")
```

We also find that are `r length(which(colSums(is.na(training)) > 0))` variables 
with missing values in the set; with a proportion of missing values as high as 
98 percent. (For the code used to quantify missing values, refer to the 
appendix.) For such variables an imputation strategy will not give reasonable
results, so we will omit them from our prediction models. As the first column is
only a record ID with no relation to the outcome, we will omit that too. This
leaves us with 58 predictor variables and the outcome.

```{r subset}
NAcols <- c(1, which(colSums(is.na(training)) > 0))
newtraining <- training[,-NAcols]
newtesting <- testing[,-NAcols]
```



***
### Model selection and training

For this exercise we use a random forests method, including cross-validation. To
do this, we use the train function from the caret package, and specify the
method as "rf". We specify the number of trees to grow as 10.

```{r train}
set.seed(54321)
fit1 <- train(classe ~., method = "rf", data = newtraining, trControl = trainControl(method = "cv"), number = 3, importance = TRUE, ntree = 10)
```



***
### Model evaluation

The resulting random forest is summarized as follows.

```{r rfmodel}
print(fit1, digits = 3)
```

By using the varImp function we can evaluate which features have the greatest importance within the model.

```{r importance}
varI <- varImp(fit1, conditional = TRUE)
plot(varI, top = 10)
```

We can evaluate the model performance by looking at the 'finalmodel' component.
The confusion matrix indicates that in the vast majority of cases, we have
correctly predicted the outcomes. The estimated out of bag error rate is 0.42%.

```{r performance}
fit1$finalModel
```



***
### Prediction

Finally, we use the predict function to predict the 20 different test cases.
```{r predict}
pred1 <- predict(fit1, newtesting)
pred1
```



***
### Appendix

####References

* Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative
Activity Recognition of Weight Lifting Exercises. Proceedings of 4th 
International Conference in Cooperation with SIGCHI (Augmented Human '13) . 
Stuttgart, Germany: ACM SIGCHI, 2013.

#### Code for quantifying missing values

* How many variables have missing values? `length(which(colSums(is.na(training)) > 0))`
* How many missing values in each variable? `sapply(training, function(x) sum(is.na(x)))`
